Higher-Order unification for free!

### 2024-06-28 paper/rebuttal.md

Dear reviewers and PC members, we are happy for the accurate and positive
reviews.
We take the occasion to answer a few questions, hopefully to satisfy the
interest of the reviewers.

----------------------- REVIEW 1 ---------------------
> Line 1368: The authors mention that one reason for delaying the
> application of substitutions is to keep the term structure indexable.
> One can imagine another reason for making such a delay, especially in
> a logic programming setting, is that failure can be encountered, and
> the work done in computing complete normal forms would be wasted.  Is
> there any evidence of that justification?

According to our experience, that is specific to the use of LP in the context
of Coq, delaying substitution (esp beta reduction) was not a crucial advantage.
An early prototype of Elpi did implement an explicit substitution calculus
but we could not observe a considerable gain (enough to justify its complexity).
Our intuition is that, in Coq, long reductions are not made of beta steps,
but rather fixpoint unfolding and ML-like pattern matching (interleaved with
betas). So just delaying beta, and sharing beta-normal forms, is not enough to
increase the performance in this context.

We hope to have correctly understood the question.

> It might interest the reader to explain what features of lambda Prolog
> that are not in Prolog remain in your specification.  For example,
> your code does not reduce HOU to first-order logic/Prolog.  The
> universal quantifier in goals and the pattern fragment unification are
> critical to your encoding, right?

We absolutely need:

1. pattern unification (and pi to induce fresh nominals)
2. delaying of goals (our links)

CHR rules, or a reified main loop (scheduling of goals), is needed only to
guarantee fidelity, that amounts at failing as soon as possible.
This may or may not be needed, depending on the use case.
It surely is in ours.

----------------------- REVIEW 2 ---------------------
> "our real goal is to simulate the execution of an entire logic-program."
> I do not understand what you mean by a "real goal".
> Is this in contrast to some goals that are not "real". I do not understand what this means.

Apologies for our English, we are not native speakers.

What we meant is that our ultimate/true goal is not (only) to implement an HO
unification procedure, but rather to simulate the execution of a whole
logic-program, that is made of many calls to that unification procedure.

We will reformulate and avoid the word "real", it is maybe a false friend with
"true" (in Italian).

> "so that the code of (𝑚𝑎𝑙𝑙𝑜𝑐) below can preserve:"
> what can it preserve?

Sorry again, sloppy wording.

We should have written "so that the code of (𝑚𝑎𝑙𝑙𝑜𝑐) can preserve
Invariant 1 written below".

----------------------- REVIEW 3 ---------------------
> Usually, I think of the eta rule as a type-directed expansion (capturing
> a form of extensionality) rather than a contraction.  I wonder if the
> authors could expand on (a) on the significance of eta for the object
> language as it relates to type-class resolution, and (b) whether it might
> be possible to make encoding sufficiently aware of the object-language
> types as eta might be modeled via an expansion.

The main reason why we did focus on contraction is that it is what the user
wants to see and typically writes when declaring type class instances. Eg
"Decidable prime" and not "Decidable (fun x => prime x)".

DAVIDE CHECK THIS: ma dove è che contraiamo? si può fare (b) facilmente? Serve chiamare coq.typecheck?

> In terms of related work, I might point out the paper
> "Higher-Order Logic Programming as Constraint Logic Programming" by
> Michaylov and Pfenning from PPCP 1993 which makes similar points as
> lines 1287-1291.  The Twelf manual also describes how nonpattern constraints
> are postponed, and how other constraint domains are integrated (Chapters 5.5 and 6),
> although not at the level of generality as CHR.

We sincerely thank the reviewer for this pointer.

### 2024-06-28 paper/rebuttal.md

We should have written "so that the code of (𝑚𝑎𝑙𝑙𝑜𝑐) can preserve
Invariant 1 written below".

----------------------- REVIEW 3 ---------------------
> Usually, I think of the eta rule as a type-directed expansion (capturing
> a form of extensionality) rather than a contraction.  I wonder if the
> authors could expand on (a) on the significance of eta for the object
> language as it relates to type-class resolution, and (b) whether it might
> be possible to make encoding sufficiently aware of the object-language
> types as eta might be modeled via an expansion.

The main reason why we did focus on contraction is that it is what the user
wants to see and typically writes when declaring type class instances. Eg
"Decidable prime" and not "Decidable (fun x => prime x)".

On the other hand, we think it is possible to produce "eta-expanded" solutions
by slightly modifying our decompilation procedure: note that the substitution we
provide to the object language is produced at the very end of the unification
procedure. Therefore, let `MV` and `OV` a mapping of a meta variable and an
object variable. If in the meta-substitution sigma, we assign a term `T` to
`MV`, then when we translate the `T` for `OV`, we can, at first, get the type of
`OV`, say `a -> b -> c`. If `T` is like `lambda x. f x x` then we can decompile
it into `lambda y z x.(f x x) z y`

### 2024-06-28 paper/rebuttal.md

Dear reviewers and PC members, we are happy for the accurate and positive
reviews.
We take the occasion to answer a few questions, hopefully to satisfy the
interest of the reviewers.

----------------------- REVIEW 1 ---------------------
> Line 1368: The authors mention that one reason for delaying the
> application of substitutions is to keep the term structure indexable.
> One can imagine another reason for making such a delay, especially in
> a logic programming setting, is that failure can be encountered, and
> the work done in computing complete normal forms would be wasted.  Is
> there any evidence of that justification?

We hope to have correctly understood the question.

In our experience type-class search does not exhibit much backtracking,
especially if the rules are indexed properly, deep enough to find the
discriminating term constructor. It is out of scope for this paper, but
another aspect we are investigating is a functionality/determinacy analysis
since many type classes happen to be functions.

@@

> It might interest the reader to explain what features of lambda Prolog
> that are not in Prolog remain in your specification.  For example,
> your code does not reduce HOU to first-order logic/Prolog.  The
> universal quantifier in goals and the pattern fragment unification are
> critical to your encoding, right?

We absolutely need:

1. pattern unification (and pi to induce fresh nominals)
2. delaying of goals (our links)

CHR rules, or a reified main loop (scheduling of goals), is needed only to
guarantee fidelity, that amounts at failing as soon as possible.
This may or may not be needed, depending on the use case.
It surely is needed for our use case.

We will try to make these points clear in the conclusion.

@@

We should have written "so that the code of (𝑚𝑎𝑙𝑙𝑜𝑐) can preserve
Invariant 1 written below".

----------------------- REVIEW 3 ---------------------
> Usually, I think of the eta rule as a type-directed expansion (capturing
> a form of extensionality) rather than a contraction.  I wonder if the
> authors could expand on (a) on the significance of eta for the object
> language as it relates to type-class resolution, and (b) whether it might
> be possible to make encoding sufficiently aware of the object-language
> types as eta might be modeled via an expansion.

The main reason why we did focus on contraction is that it is what the user
wants to see and typically writes when declaring type class instances. Eg
"Decidable prime" and not "Decidable (fun x => prime x)".

On the other hand, we think it is possible to produce "eta-expanded" solutions
by slightly modifying our decompilation procedure that is run at the very end
in order to produce the substitution for the object language.
Since Coq-Elpi runs withing Coq we could invoke Coq's type checker and use
the type to perform eta expansion.

@@

> In terms of related work, I might point out the paper
> "Higher-Order Logic Programming as Constraint Logic Programming" by
> Michaylov and Pfenning from PPCP 1993 which makes similar points as
> lines 1287-1291.  The Twelf manual also describes how nonpattern constraints
> are postponed, and how other constraint domains are integrated (Chapters 5.5 and 6),
> although not at the level of generality as CHR.

We sincerely thank the reviewer for these pointers, we will integrate them in
the last section.

### 2024-07-01 paper/TODO_rebuttal.md

# REVIEW 1

- [ ] Line 1368: The authors mention that one reason for delaying the
  application of substitutions is to keep the term structure indexable.
  One can imagine another reason for making such a delay, especially in
  a logic programming setting, is that failure can be encountered, and
  the work done in computing complete normal forms would be wasted.  Is
  there any evidence of that justification?
- [ ] It might interest the reader to explain what features of lambda Prolog
  that are not in Prolog remain in your specification.  For example,
  your code does not reduce HOU to first-order logic/Prolog.  The
  universal quantifier in goals and the pattern fragment unification are
  critical to your encoding, right?
- [ ] line 26: This is the first time I've seen the "\beta\eta"
   conversion written as "\eta\beta$.  Is there a good reason to
   switch this order from the normal?
- [ ] line 70: "for the Coq's system" - rewrite
- [ ] line 112: The reference for pattern unification is not [11] but
   Miller's 1991 paper in the J. of Logic and Computation.
- [ ] line 247: Using "subscript" as a verb is odd.  Revise that
   sentence.
- [ ] line 240: Definition 2.6 asks that a term is an eta expansion.
   Does this mean the term is an eta-redex, a term of the form 
   (lambda x. M x) where x is not free in M ?
- [ ] line 1117 "desappears"
- [ ] line 1301: "heigenvariables"
- [ ] line 1311: "mathes"
- [ ] line 1324; "since it removes all THE data from THE rule's heads"
- [ ] line 1335: "to accommodate for the one" drop "for"?
- [ ] line 1355: "reather"
- [ ] An early attempt to reduce HOU (at least, 2nd order unification) to
   the pattern fragment can be found in Miller 1991 paper "Unification
   of Simply Typed Lambda-Terms as Logic Programming", ICLP91. That
   early approach does not provide the flexibility apparent in the
   present paper.

# REVIEW 2

- [ ] The abstract is a little bit vague. It writes "more well-behaved" and 
      "problems" without any hint about what this might mean. A stronger 
      abstract would define the problem more clearly.
- [ ] "Instance forall_dec: ∀A P, Finite A → (* r3 *) ∀x:A, Decision (P x) →
      Decision (∀x:A, P x)." I think parentheses are missing around "∀x:A, 
      Decision (P x)"?
- [ ] "an Higher-Order Abstract Syntax" an => a
- [ ] "for the Coq’s system" => "for Coq's type system"
- [ ] I'm a bit surprised it is called "good unification" and not "most general 
      unification"
- [ ] "our real goal is to simulate the execution of an entire logic-program."
      I do not understand what you mean by a "real goal". Is this in contrast 
      to some goals that are not "real". I do not understand what this means.
- [ ] "so that the code of (𝑚𝑎𝑙𝑙𝑜𝑐) below can preserve" what can it preserve?
- [ ] "ensures the mapping are bijective" are => is
- [ ] In the definition of comp: "(arity z)" What is "arity z"?

# REVIEW 3

- [ ] Usually, I think of the eta rule as a type-directed expansion (capturing
    a form of extensionality) rather than a contraction.  I wonder if the
    authors could expand on (a) on the significance of eta for the object
    language as it relates to type-class resolution, and (b) whether it might
    be possible to make encoding sufficiently aware of the object-language
    types as eta might be modeled via an expansion.
- [ ] In terms of related work, I might point out the paper
    "Higher-Order Logic Programming as Constraint Logic Programming" by
    Michaylov and Pfenning from PPCP 1993 which makes similar points as
    lines 1287-1291.  The Twelf manual also describes how nonpattern constraints
    are postponed, and how other constraint domains are integrated (Chapters 5.5 and 6),
    although not at the level of generality as CHR.

### 2024-07-01 paper/TODO_rebuttal.md

# REVIEW 1

- [ ] Line 1368: The authors mention that one reason for delaying the
  application of substitutions is to keep the term structure indexable.
  One can imagine another reason for making such a delay, especially in
  a logic programming setting, is that failure can be encountered, and
  the work done in computing complete normal forms would be wasted.  Is
  there any evidence of that justification?
- [ ] It might interest the reader to explain what features of lambda Prolog
  that are not in Prolog remain in your specification.  For example,
  your code does not reduce HOU to first-order logic/Prolog.  The
  universal quantifier in goals and the pattern fragment unification are
  critical to your encoding, right?
- [ ] line 26: This is the first time I've seen the "\beta\eta"
   conversion written as "\eta\beta$.  Is there a good reason to
   switch this order from the normal?
   > Comment: We use beta-eta since we treat beta before eta in the paper
- [ ] line 70: "for the Coq's system" - rewrite
- [ ] line 112: The reference for pattern unification is not [11] but
   Miller's 1991 paper in the J. of Logic and Computation.
- [ ] line 247: Using "subscript" as a verb is odd.  Revise that
   sentence.
- [ ] line 240: Definition 2.6 asks that a term is an eta expansion.
   Does this mean the term is an eta-redex, a term of the form 
   (lambda x. M x) where x is not free in M ?
- [ ] line 1117 "desappears"
- [ ] line 1301: "heigenvariables"
- [ ] line 1311: "mathes"
- [ ] line 1324; "since it removes all THE data from THE rule's heads"
- [ ] line 1335: "to accommodate for the one" drop "for"?
- [ ] line 1355: "reather"
- [ ] An early attempt to reduce HOU (at least, 2nd order unification) to
   the pattern fragment can be found in Miller 1991 paper "Unification
   of Simply Typed Lambda-Terms as Logic Programming", ICLP91. That
   early approach does not provide the flexibility apparent in the
   present paper.

### 2024-07-01 paper/TODO_rebuttal.md

# REVIEW 1

- [ ] Line 1368: The authors mention that one reason for delaying the
  application of substitutions is to keep the term structure indexable.
  One can imagine another reason for making such a delay, especially in
  a logic programming setting, is that failure can be encountered, and
  the work done in computing complete normal forms would be wasted.  Is
  there any evidence of that justification?
- [ ] It might interest the reader to explain what features of lambda Prolog
  that are not in Prolog remain in your specification.  For example,
  your code does not reduce HOU to first-order logic/Prolog.  The
  universal quantifier in goals and the pattern fragment unification are
  critical to your encoding, right?
- [ ] line 26: This is the first time I've seen the "\beta\eta"
   conversion written as "\eta\beta$.  Is there a good reason to
   switch this order from the normal?
   > Comment: We use beta-eta since we treat beta before eta in the paper
- [ ] line 70: "for the Coq's system" - rewrite
- [x] line 112: The reference for pattern unification is not [11] but
   Miller's 1991 paper in the J. of Logic and Computation. (cf 827b11dd2df213b9f446b77434fef8cb6591f00b)
- [ ] line 247: Using "subscript" as a verb is odd.  Revise that
   sentence.
- [ ] line 240: Definition 2.6 asks that a term is an eta expansion.
   Does this mean the term is an eta-redex, a term of the form 
   (lambda x. M x) where x is not free in M ?
- [ ] line 1117 "desappears"
- [ ] line 1301: "heigenvariables"
- [ ] line 1311: "mathes"
- [ ] line 1324; "since it removes all THE data from THE rule's heads"
- [ ] line 1335: "to accommodate for the one" drop "for"?
- [ ] line 1355: "reather"
- [ ] An early attempt to reduce HOU (at least, 2nd order unification) to
   the pattern fragment can be found in Miller 1991 paper "Unification
   of Simply Typed Lambda-Terms as Logic Programming", ICLP91. That
   early approach does not provide the flexibility apparent in the
   present paper.

### 2024-07-01 paper/main.tex
correct ref to miller
desappears -> disappears
mathes -> matches

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second-order parameter \coqIn{P} that represents
a function of type \coqIn{A ~$\to$~ Prop} (a 1-argument predicate over \coqIn{A}).
The solver infers a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular, it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
pattern fragment and admits a unique solution:
\begin{elpicode}
P = ~$\lambda$~x.nfact x 3                                       ~\customlabel{solution:intro}{(\rho)}~
\end{elpicode}
In the rest of this paper, we refer to the pattern fragment as
\llambda~\cite{miller92jsc}.

@@

\noindent
Note that $\mapping{X}{C}{0}$ desappears from \mapStore
in favour of the auxiliary \linketa \linketaM{}{C}{\lambda x.B_{x}}.
The resolution of $\hoUnifPb_1$ assignes $a$ to $A$. This wakes up
$\linkStore_2$ by \progressetaleft, assigning <<$f \appsep x$>> to $B_x$.
The resolution of $\hoUnifPb_2$ instantiates $C$ to $\lambda x.a$, which, in
turn triggers $\linkStore_1$ by \progressetaleft. In turn, this performs
$\lambda x.a \Ue \lambda x.(f \appsep x)$ that fails as expected.

@@

\noindent
The first directive matches a constraint
whose \lhs is a variable \elpiIn{X}; the second mathes and removes
a constraint on the same variable; the third
relocates \elpiIn{T2} and the latter crafts the new goal that unifies
\elpiIn{T1} with \elpiIn{T2'} under \elpiIn{G1} and the set of
heigenvariables \elpiIn{N1}.

>>>>

\noindent
The proof found by the solver back-chains on rule 3 (the only rule
about the $\forall$ quantifier), and then solves the premises with
rules 1 and 2 respectively.
Note that rule 3 features a second-order parameter \coqIn{P} that represents
a function of type \coqIn{A ~$\to$~ Prop} (a 1-argument predicate over \coqIn{A}).
The solver infers a value for \coqIn{P} by unifying the conclusion
of rule 3 with the goal, and in particular, it has to solve the unification
problem \coqIn{P x = nfact x 3}. This higher-order problem falls in the so-called 
pattern fragment and admits a unique solution:
\begin{elpicode}
P = ~$\lambda$~x.nfact x 3                                       ~\customlabel{solution:intro}{(\rho)}~
\end{elpicode}
In the rest of this paper, we refer to the pattern fragment as
\llambda~\cite{miller1991}.

@@

\noindent
Note that $\mapping{X}{C}{0}$ disappears from \mapStore
in favour of the auxiliary \linketa \linketaM{}{C}{\lambda x.B_{x}}.
The resolution of $\hoUnifPb_1$ assignes $a$ to $A$. This wakes up
$\linkStore_2$ by \progressetaleft, assigning <<$f \appsep x$>> to $B_x$.
The resolution of $\hoUnifPb_2$ instantiates $C$ to $\lambda x.a$, which, in
turn triggers $\linkStore_1$ by \progressetaleft. In turn, this performs
$\lambda x.a \Ue \lambda x.(f \appsep x)$ that fails as expected.

@@

\noindent
The first directive matches a constraint
whose \lhs is a variable \elpiIn{X}; the second matches and removes
a constraint on the same variable; the third
relocates \elpiIn{T2} and the latter crafts the new goal that unifies
\elpiIn{T1} with \elpiIn{T2'} under \elpiIn{G1} and the set of
heigenvariables \elpiIn{N1}.
